Spike spectra for recurrences

### 2022-07-27 Manuscript/inter_spike_manuscript_mdpi.tex
Laskar section added

We exemplify the use of the inter spike spectrum in combination with the $\tau$-RR as outlined in 
Section \ref{sec_tau_rr_intro} on several interesting research questions. The procedure is the following:
\begin{itemize}[noitemsep]
\item[(1)] Compute a RP of the trajectory of the system, Eq.~\eqref{eq_rp_definition}. 
\item[(2)] Compute the $\tau$-RR of that RP, Eq.~\eqref{eq_tau_rr}.
\item[(3)] Transform the $\tau$-RR into the proposed inter spike spectrum, see Section \ref{sec_tau_rr_method}.
\end{itemize}

@@

The Fourier frequency spectra in Figure~\ref{fig_power_grid_spectra}A,B, however, do not display sharp peaks exactly at $15$, $30$ and $60$ \si{min} (for technical details on the calculation of 
the spectra shown, the reader is referred to the Appendix~\ref{sec_power_grid_appendix}). Especially the $30$ and $60$ \si{min} 
peaks are split into two adjacent peaks and the local minimum in between these ``double''-peaks correspond to the exact times. 
In contrast, the $15$ \si{min} peak is completely absent from the inter-spike spectra of the $\tau$-RRs of the frequency data (Fig.~\ref{fig_power_grid_spectra}C,D), but these show sharp peaks 
at $30$ and $60$ \si{min}. In front of these peaks, smaller local peaks can be seen, which correspond to the local peaks of the Fourier power spectra at $28$ and $55$ \si{min}, respectively.
Moreover, there is an additional peak at 
$40$ \si{min} for both datasets, which is absent in the Fourier spectrum and which is not a multiple of the missing $15$ \si{min} peak. The position and magnitude of the peaks in the shown 
inter spike spectra are robust to the chosen recurrence threshold, the regression method and its regularization as well as the sampling time of the original signal.\\

We interpret the results presented as follows. The missing $15$ \si{min} in the inter spike spectra is due to the much stronger autocorrelation at $30$ and $60$ \si{min} 
(see Fig.~\ref{fig_power_grid_time_series}B,D) and because these periods are integer multiples of $15$ \si{min} the inter spike spectra are not able to detect it and the 
sparse regression ``drags'' the $15$ \si{min} periods into the mentioned $30$ and $60$ \si{min} peaks. Moreover, the inter spike spectra, unlike the Fourier spectra demonstrates sharp peaks exactly at $30$ and $60$ \si{min}, i.e. during dispatches. Eventually, here we have found a clear sharp peak at $40$ \si{min} which can occur because of any regular controls in a power grid, and can be a hint to develop the existing stochastic processes to model precisely the power grid frequency \cite{gorjao2020data}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Evolutionary inter spike spectra of astronomical data}

\todo[inline]{Add this section and keep it very very short.}

>>>>

We exemplify the use of the inter spike spectrum in combination with the $\tau$-RR as outlined in 
Section \ref{sec_tau_rr_intro} on several interesting research questions. The procedure is the following:
\begin{itemize}[noitemsep]
\item[(1)] Compute a RP of the trajectory of the system, Eq.~\eqref{eq_rp_definition}. If only univariate data is available, perform a state space reconstruction for obtaining the trajectory first.
\item[(2)] Compute the $\tau$-RR of that RP, Eq.~\eqref{eq_tau_rr}.
\item[(3)] Transform the $\tau$-RR into the proposed inter spike spectrum, see Section \ref{sec_tau_rr_method}.
\end{itemize}

@@

The Fourier frequency spectra for the Central Europe data in Figure~\ref{fig_power_grid_spectra}A, however, does not display sharp peaks exactly at $15$, $30$ and $60$ \si{min} 
(for technical details on the calculation of 
the spectra shown, the reader is referred to the Appendix~\ref{sec_power_grid_appendix}). Especially the $30$ and $60$ \si{min} 
peaks are split into two adjacent peaks and the local minimum in between these ``double''-peaks correspond to the exact times. The Great Britain counterpart in Fig.~\ref{fig_power_grid_spectra}B 
has sharp peaks at $15$ and $30$ \si{min} and also ``local predecessor peaks'' for $30$ and $60$ \si{min} at the same positions as in panel A. 
In contrast, the $15$ \si{min} peak is completely absent from the inter-spike spectra of the $\tau$-RRs of the frequency data (Fig.~\ref{fig_power_grid_spectra}C,D), but these show sharp peaks 
at $30$ and $60$ \si{min}. In front of these peaks, smaller local peaks can be seen, which correspond to the local peaks of the Fourier power spectra at $28$ and $55$ \si{min}, respectively.
Moreover, there is an additional peak at 
$40$ \si{min} for both datasets, which is absent in the Fourier spectrum and which is not a multiple of the missing $15$ \si{min} peak. The position and magnitude of the peaks in the shown 
inter spike spectra are robust to the chosen recurrence threshold, the regression method and its regularization as well as the sampling time of the original signal.\\

We interpret the results presented as follows. The missing $15$ \si{min} in the inter spike spectra is due to the much stronger autocorrelation at $30$ and $60$ \si{min} 
(see Fig.~\ref{fig_power_grid_time_series}B,D) and because these periods are integer multiples of $15$ \si{min} the inter spike spectra are not able to detect it and the 
sparse regression ``drags'' the $15$ \si{min} periods into the mentioned $30$ and $60$ \si{min} peaks. Moreover, the inter spike spectra, unlike the Fourier spectra demonstrates sharp peaks exactly 
at $30$ and $60$ \si{min}, i.e. during dispatches (not valid in case of the GB dataset). Eventually, here we have found a clear sharp peak at $40$ \si{min} which can occur 
because of any regular controls in a power grid, and can be a hint to develop the existing stochastic processes to model precisely the power grid frequency \cite{gorjao2020data}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Evolutionary inter spike spectra of Earth's orbit data}

When applying the proposed inter spike spectrum to the $\tau$-RR of a time series we expect additional frequency/period information, due to the fact that the recurrence plot (RP), 
Eq.~\eqref{eq_rp_definition}, and its corresponding $\tau$-RR, Eq.~\eqref{eq_tau_rr}, visualize the trajectory of the embedded time series in an embedding space of higher dimension. 
However, given a sufficient embedding of the time series, we would also expect that major frequencies/periods of the non-embedded time series are incorporated in the RP, its $\tau$-RR, 
and eventually in the inter spike spectrum of the $\tau$-RR. In order to demonstrate this we apply the inter spike spectrum to the freely available eccentricity time series of \citet{Laskar2011}. 
This astronomical computation of the orbital motion of the Earth (here we focus on the eccentricity only) has a clear expectation value of the incorporated frequencies/periods. The three leading 
eccentricity cycles of 405 kyr period, 95 kyr period and 124 kyr period are well known in paleoclimate studies \cite{Laskar2004}. Our aim in this section is to show that the inter spike spectrum 
of the $\tau$-RR of the embedded eccentricity time series will reflect these cycles in a similar fashion as the Fourier powerspectrum of the non-embedded eccentricity time series. We will further 
show that Fourier-transforming the $\tau$-RR instead of applying the proposed inter spike spectrum will lead to non satisfying results, since the spiky $\tau$-RR excites a variety of harmonics in 
the corresponding Fourier spectrum (cf. Section \ref{sec_tau_rr_intro}). We use a time series which covers the past \mytilde $67$ Mio. years, which a total length of $N=13,421$ samples and a 
sampling period of $\Delta t= 5,000$.\\
First we compute an evolutionary short time Fourier transform (FT) using a windowsize of $ws=1,000$ samplepoints ($\equiv 5 Mys$) shifted by unity and a 
Hamming window. The spectrogram is shown in Figure \ref{fig_laskar_spectra}A, where the expected periods mentioned and highlighted are clearly visible. 
The spectrogram is shown in Figure A, where the expected periods mentioned and highlighted are clearly visible.

\begin{figure}
 \centering
 \includegraphics[width=\textwidth]{./figures/fig_laskar_spectra}
 \caption{\textbf{A} Evolutionary Fourier power spectra of eccentricity time series. \textbf{B} Inter spike spectrogram of the $\tau$-recurrence rate of the eccentricity time 
 series and \textit{C} its Fourier spectrogram. Horizontal black dashed lines highlight the analytically expected orbital periods of $405$, $124$ and $95$ kyrs. 
 For comparability, in all cases the spectra aligned in the columns of the shown plots are normalized to probabilities (sum of unity for 
 each power spectrum). For further computational details, please refer to the main text.}  
\label{fig_laskar_spectra}
\end{figure}

\noindent Then we construct the inter spike spectrogram of the $\tau$-RR by first determining an appropriate embedding. By using a recent Tree-Embedding-Ansatz \cite{Kraemer2022} we minimize the 
\textit{false nearest neighbor} statistic \cite{hegger1999} and use the continuity statistic \cite{pecora2007} for potential delays. Eventually we obtain an $8$-dimensional embedding 
with delays $\tau s=[0, 18, 36, 48, 62, 101, 114]$ (in sampling units) for the entire time series. Similar to the preceding approach we embed the time series with these embedding parameters 
on windows of size $ws=1,200$ and a unity shift. \footnote{We use a slightly larger window here than in the short FT, because the embedding causes a ``loss'' of data points and we want to 
cover similar time spans.} The RPs are computed on these embedded trajectories with a fixed recurrence threshold corresponding to $10\%$ global recurrence rate \cite{kraemer2018} 
and the inter spike spectra of the corresponding $\tau$-RR's is obtained by using STLS regression (see Section \ref{sec_tau_rr_method}) and a regularization threshold corresponding to 
$\rho=0.9$ accordance of $\tau$-RR's and re-composed signals. We only use the first $200$ data points of the $\tau$-RR (covering a time span of $1$ Myrs). 
The spectrogram shown in Figure \ref{fig_laskar_spectra}B also highlights the $95$, $124$ and $405$ kyrs periods as expected. 
Additional power is distributed in its harmonics at $190$ and $248$ kyrs periods. Finally, the standard FT with a Hamming window of the same $\tau$-RR's used to obtain the inter spike 
spectrogram in panel B, yields a smeared spectrogram which does not reflect the expected periods, but rather suffers from the spike train behavior, i.e. many excited harmonics, of the 
FT described in Section \ref{sec_tau_rr_intro} and Fig.~\ref{fig_tau_rr_dirac_comb}. The shown results are robust to a change of embedding parameters and windowsizes. However, a too low 
regularization threshold smears the clear spectrogram in panel B.
